Clustering time series based on probability distributions across temporal granularities

### 2021-11-17 main.Rmd
fig 4 caption

```{r sindex-plot-validation, fig.cap = "The cluster separation index (sindex) is plotted as a function of the number of clusters for the range of mean differences (rows) under the different scenarios (columns). S1 has a sharp decrease in sindex from 5 to 6, whereas S2 and S3 have a decrease from 4 to 5. As a result, the number of clusters considered for S1, S2, and S3 is 5, 4, 4, respectively. This corresponds to the number of designs taken into account in each scenario."}
```

@@

Clustering with the new distances is illustrated on the smart meter energy usage for a sample of customers from [SGSC consumer trial data](https://data.gov.au/data/dataset/smart-grid-smart-city-customer-trial-data) which was available through [Department of the Environment and Energy](https://data.gov.au/data/organization/doee) and Data61 CSIRO. The full data contains half-hourly general supply in KwH for 13,735 customers, resulting in 344,518,791 observations in total.<!--In most cases, electricity data is expected to have multiple seasonal patterns like daily, weekly or annual. We do not learn about these repetitive behaviors from the linear view because too many measurements all squeezed in that representation. Hence we transition into looking at cyclic granularities, that can potentially provide more insight on their repetitive behavior.--> The raw data for these consumers is of unequal length, with varying starting and end dates.  Additionally, there were missing values in many series. (The supplementary material contains details from checking for systematic missingness.) Because our proposed methods evaluate probability distributions rather than raw data, these data issues are not problematic, unless any systematic structure unrelated to granularities.

Huge data sets presents more complications for clustering. Clustering algorithms work well when there well-separated clusters, with no nuisance variables or nuisance observations. When converting a series to granularities, many variables (each level of a granularity) are generated, possibly creating a slew of nuisance variables. Some households may have a mix of energy use patterns, which could be considered nuisance observations, between major clusters. For this reason, we have chosen to select a small group of households with relatively distinct and different patterns in order to illustrate the clustering, simply. This is often a good approach to tackling a big analysis task, so start with a simpler task. The approach, however, is applicable using on all households, but it would be more complicated to make final choice of the number of clusters and explain the results.

As a result, we dissect the larger problem and test our solutions on a small sample of prototype customers. To do this, data is first filtered to generate a clean sample, and then significant cyclic granularities (variables) for them are chosen. This is described in Section \ref{sec:datafilter}. The clean set is subsequently subsampled along all dimensions of interest, to ensure that the sampled data reveals some patterns across at least one specified variable, which is described in Section \ref{sec:prototype}. By grouping the prototypes using our methods in Section \ref{sec:clustering} and assessing their meaning, the study hopes to unravel some of the heterogeneities observed in energy usage data. Because our application does not employ additional customer data, we cannot explain why consumption varies, but rather try to identify how it varies.

@@

- Choose a smaller subset of randomly selected $600$ customers with no implicit missing values for 2013.

- Obtain $wpd$ for all cyclic granularities considered for these customers. It was found that $hod$ (hour-of-day), $moy$ (month-of-year) and $wnwd$ (weekend/weekday) are turning out to be significant for most customers. We use these three granularities while clustering.

- Remove customers whose data for an entire category of $hod$, $moy$ or $wnwd$ is empty. For example, a customer who does not have data for an entire month is excluded because their monthly behavior cannot be analyzed.

- Remove customers whose energy consumption is 0 in all deciles. These are the clients whose consumption is likely to remain essentially flat and with no intriguing repeated patterns that we are interested in studying.

## Prototype selection {#sec:prototype}

<!-- Why instance selection -->
Supervised learning uses a training set of known information to categorize new events through instance selection. Instance selection [@olvera2010review] is a method of rejecting instances that are not helpful for classification. This is analogous to subsampling the population along all dimensions of interest such that the sampled data represents the primary features of the underlying distribution. Instance selection in unsupervised learning has received little attention in the literature. However, it could be a useful tool for evaluating model or method performance. There are several ways to approach prototype selection. Following the idea of @Fan2021-bq of picking related examples (neighbors) for each instance (anchor), we can first use any dimensionality reduction techniques like MDS or PCA to project the data into a 2D space. Then pick a few "anchor" customers far apart in 2D space and pick a few neighbors for each. Unfortunately, this does not ensure that consumers with significant patterns across all variables are chosen. Tours can reveal variable separation that gets hidden in a single variable display better than static projections. Hence, we perform a linked tour with a t-SNE layout using the R package `liminal` [@R-liminal] to identify customers who are more likely to have distinct patterns across the variables studied. (Refer to the supplementary article for further details). Figure \ref{fig:hod-ind-group-png} shows the distribution across $hod$, $moy$ and $wnwd$ for the set of chosen $24$ customers that were chosen. The customers are split into batches of $12$ in (a) and (b). Each row in (a) and that in (b) represents the profile of one customer across different variables. <!--Few of these customers have similar distribution across $moy$ and some are similar in their $hod$ distribution. -->

>>>>

```{r sindex-plot-validation, fig.cap = "Choosing optimal cluster number across the range of scenarios and mean differences used in the validation study, using the cluster separation index (sindex) for the JS-NQT. S1 has a sharp decrease in sindex from 5 to 6, whereas S2 and S3 have a decrease from 4 to 5, especially when mean difference is large, providing the recommended number of clusters to be 5, 4, 4, respectively. This precisely reflects the structure in designs that we would hope the clustering could recover."}
```

@@

Clustering with the new distances is illustrated on the smart meter energy usage for a sample of customers from [SGSC consumer trial data](https://data.gov.au/data/dataset/smart-grid-smart-city-customer-trial-data) which was available through [Department of the Environment and Energy](https://data.gov.au/data/organization/doee) and Data61 CSIRO [@make-these-two-URLs-into-proper-citations]. The full data contains half-hourly general supply in KwH for 13,735 customers, resulting in 344,518,791 observations in total.<!--In most cases, electricity data is expected to have multiple seasonal patterns like daily, weekly or annual. We do not learn about these repetitive behaviors from the linear view because too many measurements all squeezed in that representation. Hence we transition into looking at cyclic granularities, that can potentially provide more insight on their repetitive behavior.--> The raw data for these consumers is of unequal length, with varying starting and end dates.  Additionally, there were missing values in many series. (The supplementary material contains details from checking for systematic missingness.) Because our proposed methods evaluate probability distributions rather than raw data, these data issues are not problematic, unless there is any systematic structure is related to granularities.

Huge data sets present more complications for clustering. Clustering algorithms work well when there are well-separated clusters, with no nuisance variables or nuisance observations. When converting a series to granularities, many variables (each level of a granularity) are generated, possibly creating a slew of nuisance variables. Some households may have a mix of energy use patterns, which could be considered nuisance observations located between major clusters. For this reason, we have chosen to select a small group of households with relatively distinct and different patterns in order to illustrate the clustering more simply. This is often a good approach to tackling a big analysis task, to start with a simpler task. The approach, however, is applicable to all households.

As a result, we dissect the larger problem and test our solutions on a small sample of prototype customers. To do this, data is first filtered to generate a small sample, and then significant cyclic granularities (variables) for them are chosen (as described in Section \ref{sec:datafilter}). The sample set is subsequently examined along all dimensions of interest, to ensure that they reveal some patterns across at least one specified variable (as described in Section \ref{sec:prototype}). <!--By grouping the prototypes using our methods in Section \ref{sec:clustering} and assessing their meaning, the study hopes to unravel some of the heterogeneities observed in energy usage data.--> Because the data does not contain additional customer characteristics, we cannot explain why consumption varies, but can only identify how it varies.

@@

The steps for household filtering and variable selection were:

1. Choose a smaller subset of randomly selected $600$ customers with no implicit missing values for 2013.

2. Obtain $wpd$ for all cyclic granularities considered for these customers. It was found that $hod$ (hour-of-day), $moy$ (month-of-year) and $wnwd$ (weekend/weekday) are turning out to be significant for most customers. We use these three granularities while clustering.

3. Remove customers whose data for an entire category of $hod$, $moy$ or $wnwd$ is empty. For example, a customer who does not have data for an entire month is excluded because their monthly behavior cannot be analyzed.

4. Remove customers whose energy consumption is 0 in all deciles. These are the clients whose consumption is likely to remain essentially flat and with no intriguing repeated patterns that we are interested in studying.

## Choosing households {#sec:prototype}

<!-- Why instance selection -->
Figure \ref{fig:hod-ind-group-png} shows the distribution across $hod$, $moy$ and $wnwd$ for the set of $24$ customers is the sample to illustrate clustering. The customers are displayed in two columns of $12$ for space reasons. Each row of each column represents the profile of a single customer across different variables.

Supervised learning uses a training set of known information to categorize new events through instance selection. Instance selection [@olvera2010review] is a method of rejecting instances that are not helpful for classification. This is analogous to subsampling the population along all dimensions of interest such that the sampled data represents the primary features of the underlying distribution. Instance selection in unsupervised learning has received little attention in the literature. However, it could be a useful tool for evaluating model or method performance. There are several ways to approach prototype selection. Following the idea of @Fan2021-bq of picking related examples (neighbors) for each instance (anchor), we can first use any dimensionality reduction techniques like MDS or PCA to project the data into a 2D space. Then pick a few "anchor" customers far apart in 2D space and pick a few neighbors for each. Unfortunately, this does not ensure that consumers with significant patterns across all variables are chosen. Tours can reveal variable separation that gets hidden in a single variable display better than static projections. Hence, we perform a linked tour with a t-SNE layout using the R package `liminal` [@R-liminal] to identify customers who are more likely to have distinct patterns across the variables studied. (Refer to the supplementary article for further details). <!--Few of these customers have similar distribution across $moy$ and some are similar in their $hod$ distribution. -->
